---
title: "Census Data Dashboard"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
runtime: shiny
---

```{r global, include=FALSE, echo = FALSE}
library(flexdashboard)
library(ggplot2)
library(tidyverse)
library(here)
library(sf)

library(tidycensus)
options(tigris_use_cache = TRUE)

library(tigris)
library(tmap)

census_key <- Sys.getenv("CENSUS_API_KEY")
knitr::opts_chunk$set(fig.width = 5, fig.asp = 1/3)
```

```{r total population 2010, echo = FALSE}
if (!file.exists(here("data", "total_pop_10.RDS"))) {
  total_population_10 <- get_decennial(
  geography = "state", 
  variables = "P001001",
  year = 2010
  )
  
  saveRDS(total_population_10, file=here("data", "total_pop_10.RDS"))
}
```

```{r total population 2020, echo = FALSE}
if (!file.exists(here("data", "total_pop_20.RDS"))) {
  total_population_20 <- get_decennial(
  geography = "state", 
  variables = "P1_001N",
  year = 2020
  )
  
  saveRDS(total_population_20, file=here("data", "total_pop_20.RDS"))
}
```

```{r Black population data 2010, echo = FALSE}
if (!file.exists(here("data", "black_pop_10.RDS"))) {
  black_pop_10 <- get_decennial(
    geography = "state",
    variables = "P001004", # Black, not Hispanic or Latino
    year = 2010,
    sumfile = "pl",
    summary_var = "P001001",
    geometry = TRUE
  )
  saveRDS(black_pop_10, file=here("data", "black_pop_10.RDS"))
}
```

```{r Black population data 2020, echo = FALSE}
if (!file.exists(here("data", "black_pop_20.RDS"))) {
  black_pop_20 <- get_decennial(
  geography = "state", 
  variables = "P1_004N", # Black, not Hispanic or Latino
  year = 2020,
  sumfile = "pl",
  summary_var = "P1_001N",
  geometry = TRUE
  )
  
  saveRDS(black_pop_20, file=here("data", "black_pop_20.RDS"))
}
```

```{r DC Black data from 2020, echo = FALSE}
if (!file.exists(here("data", "black_dc_20.RDS"))) {
  black_dc_20 <- get_decennial(
  geography = "tract", 
  state = "DC",
  variables = "P1_004N", # Black, not Hispanic or Latino
  year = 2020,
  sumfile = "pl",
  summary_var = "P1_001N",
  geometry = TRUE
  )
  
  saveRDS(black_dc_20, file=here("data", "black_dc_20.RDS"))
}
```

```{r ND Black data from 2020, echo = FALSE}
if (!file.exists(here("data", "black_nd_20.RDS"))) {
  black_nd_20 <- get_decennial(
  geography = "tract", 
  state = "ND",
  variables = "P1_004N", # Black, not Hispanic or Latino
  year = 2020,
  sumfile = "pl",
  summary_var = "P1_001N",
  geometry = TRUE
  )
  
  saveRDS(black_nd_20, file=here("data", "black_nd_20.RDS"))
}
```

```{r read in data from RDS files, echo = FALSE}
black_10 <- as_tibble(readRDS(here("data", "black_pop_10.RDS")))
black_20 <- as_tibble(readRDS(here("data", "black_pop_20.RDS")))
total_pop_10 <- as_tibble(readRDS(here("data", "total_pop_10.RDS")))
total_pop_20 <- as_tibble(readRDS(here("data", "total_pop_20.RDS")))
black_dc_20 <- as_tibble(readRDS(here("data", "black_dc_20.RDS")))
black_nd_20 <- as_tibble(readRDS(here("data", "black_nd_20.RDS")))
```

# About

# The Data

# Static 1

## Column 1

# Static 2

## Row

# Interactive 1

# Interactive 2

# Analysis

### Plotting total population by state for 2010

```{r plotting total population by state for 2010}
total_pop_10 %>% 
  arrange(desc(value)) %>%
  mutate(ten_thou_pop = map_dbl(value, function(x) { return(x/10000) }) ) %>%
  ggplot(aes(x = ten_thou_pop, y = reorder(NAME, ten_thou_pop))) +
  geom_point() +
  labs(
    title = "Total population by state from the 2010 US Census",
    subtitle = "Arranged in descending order of population",
    x = "Population (in ten thousands)",
    y = "US State"
  )
```

### Plotting total population by state for 2020

```{r plotting total population by state for 2020}
total_pop_20 %>% 
  arrange(desc(value)) %>%
  mutate(ten_thou_pop = map_dbl(value, function(x) { return(x/10000) }) ) %>%
  ggplot(aes(x = ten_thou_pop, y = reorder(NAME, ten_thou_pop))) +
  geom_point() +
  labs(
    title = "Total population by state from the 2020 US Census",
    subtitle = "Arranged in descending order of population",
    x = "Population (in ten thousands)",
    y = "US State"
  )
```

### Visualizing Black population as percentage of total population in 2020

```{r visualizing Black population as percentage of total population in 2020}
p <- st_sf(black_20) %>%
  mutate(percent = 100 * (value / summary_value)) %>%
  ggplot(aes(fill = percent)) +
  geom_sf() +
  scale_fill_distiller(palette = "RdPu", 
                       direction = 1) +
  labs(title = "Black population by State, 2020",
       caption = "Data source: 2020 1-year ACS, US Census Bureau",
       fill = "Percentage of population that is Black") + 
  theme_void()

p
```

### Visualizing Black population in DC

DC is the state with the highest percentage Black population in 2020, so I'm interested to see how that plays out within the state.

```{r visualize Black populations in DC}
dc_black_pop_20 <- st_sf(black_dc_20) %>%
  mutate(percent = 100 * (value / summary_value))
  
tm_shape(dc_black_pop_20) +
  tm_polygons(col = "percent",
          style = "quantile",
          n = 5,
          palette = "Purples",
          title = "2020 US Census") + 
  tm_layout(title = "Percent Black\nby Census tract in DC",
            frame = FALSE,
            legend.outside = TRUE)
```

## Analyzing change in Black population since 2010

```{r analyzing change in Black population since 2010}
black_10_clean <- black_10 %>%
  st_drop_geometry() %>%
  select(GEOID, value, summary_value) %>%
  rename(value10 = value) %>%
  rename(summary10 = summary_value)

black_20_clean <- black_20 %>%
  rename(value20 = value) %>%
  rename(summary20 = summary_value)

black_joined <- black_20_clean %>%
  left_join(black_10_clean, by = "GEOID")

black_change <- black_joined %>%
  mutate(
    total_change = value20 - value10,
    percent_change = 100 * (total_change / value10)
  )
```

### Visualizing percentage change for Black population

```{r visualizing percentage change for Black population}
black_change %>%
  ggplot(aes(x = percent_change, y = reorder(NAME, percent_change))) +
  geom_point() +
  labs(
    title = "Visualizing percentage change for Black population from the 2010 and 2020 US census data",
    subtitle = "Arranged in descending order of percentage change",
    x = "Percent change from 2010 to 2020",
    y = "State"
  )
```

From this graph, we see that the state with the most Black population growth is North Dakota with an almost 250% increase, while Puerto Rico has actually had a negative decline in Black population.

### Visualizing Black population in ND

DC is the state with the most Black population growth in 2020, so I'm interested to see how that plays out within the state.

```{r}
nd_black_pop_20 <- st_sf(black_nd_20) %>%
  mutate(percent = 100 * (value / summary_value))
  
tm_shape(nd_black_pop_20) +
  tm_polygons(col = "percent",
          style = "quantile",
          n = 5,
          palette = "Purples",
          title = "2020 US Census") + 
  tm_layout(title = "Percent Black\nby Census tract in ND",
            frame = FALSE,
            legend.outside = TRUE)
```

### Compare state with highest percentage of Black population (DC) with state with most Black population growth (ND)

```{r facet DC vs ND}
black_pop_10 <- black_10 %>%
  mutate(year = 2010)

black_pop_20 <- black_20 %>%
  mutate(year = 2020)

black_dc_nd <- black_pop_10 %>%
  bind_rows(black_pop_20) %>%
  rename(state = "NAME") %>%
  mutate(percent_black = value * 100 /summary_value) %>%
  mutate(year = as.integer(year)) %>%
  mutate(state = fct(state)) %>%
  filter(state == "North Dakota" | state == "District of Columbia") %>%
  group_by(state)

black_dc_nd %>%
  ggplot(aes(year, percent_black, fill = state)) +
  geom_bar(stat = "identity") +
  geom_line() +
  geom_point() +
  facet_grid(rows = vars(state)) +
  labs(
    title = "Comparing highest Black population with most Black population growth",
    subtitle = "Looking at DC and ND from US Census Data 2010-2020",
    x = "Year",
    y = "Percentage of population that is Black"
  )
```

This last graph is just to satisfy my curiosity. It's interesting to see how the state that had the most Black population growth doesn't have 5% Black people even in 2020. And though DC's Black population actually decreased percentage-wise, they still have at least 4x more Black people.