---
title: "Census Data Dashboard"
output:
  flexdashboard::flex_dashboard:
    storyboard: true
    vertical_layout: scroll
runtime: shiny
resource_files:
- data/medinc_race.RDS
- data/black_pop_5_yr.RDS
- data/pa_county_20.RDS
- data/fl_county_20.RDS
- data/ca_county_20.RDS
- data/black_nd_20.RDS
- data/black_dc_20.RDS
- data/black_pop_20.RDS
- data/black_pop_10.RDS
- data/total_pop_20.RDS
- data/total_pop_10.RDS
- data/medinc_race_long.RDS
---

```{r global, include=FALSE, echo = FALSE}
library(flexdashboard)
library(ggplot2)
library(tidyverse)
library(here)
library(shiny)
library(sf)

library(tidycensus)
options(tigris_use_cache = TRUE,
        scipen = 100)

library(tigris)
library(tmap)

census_key <- Sys.getenv("CENSUS_API_KEY")
knitr::opts_chunk$set(fig.width = 5, fig.asp = 1/3)
```

```{r total population 2010, echo = FALSE}
if (!file.exists(here("data", "total_pop_10.RDS"))) {
  total_population_10 <- get_decennial(
  geography = "state", 
  variables = "P001001",
  year = 2010
  )
  
  saveRDS(total_population_10, file=here("data", "total_pop_10.RDS"))
}
```

```{r total population 2020, echo = FALSE}
if (!file.exists(here("data", "total_pop_20.RDS"))) {
  total_population_20 <- get_decennial(
  geography = "state", 
  variables = "P1_001N",
  year = 2020
  )
  
  saveRDS(total_population_20, file=here("data", "total_pop_20.RDS"))
}
```

```{r Black population data 2010, echo = FALSE}
if (!file.exists(here("data", "black_pop_10.RDS"))) {
  black_pop_10 <- get_decennial(
    geography = "state",
    variables = "P001004", # Black, not Hispanic or Latino
    year = 2010,
    sumfile = "pl",
    summary_var = "P001001",
    geometry = TRUE
  )
  saveRDS(black_pop_10, file=here("data", "black_pop_10.RDS"))
}
```

```{r Black population data 2020, echo = FALSE}
if (!file.exists(here("data", "black_pop_20.RDS"))) {
  black_pop_20 <- get_decennial(
  geography = "state", 
  variables = "P1_004N", # Black, not Hispanic or Latino
  year = 2020,
  sumfile = "pl",
  summary_var = "P1_001N",
  geometry = TRUE
  )
  
  saveRDS(black_pop_20, file=here("data", "black_pop_20.RDS"))
}
```

```{r DC Black data from 2020, echo = FALSE}
if (!file.exists(here("data", "black_dc_20.RDS"))) {
  black_dc_20 <- get_decennial(
  geography = "tract", 
  state = "DC",
  variables = "P1_004N", # Black, not Hispanic or Latino
  year = 2020,
  sumfile = "pl",
  summary_var = "P1_001N",
  geometry = TRUE
  )
  
  saveRDS(black_dc_20, file=here("data", "black_dc_20.RDS"))
}
```

```{r ND Black data from 2020, echo = FALSE}
if (!file.exists(here("data", "black_nd_20.RDS"))) {
  black_nd_20 <- get_decennial(
  geography = "tract", 
  state = "ND",
  variables = "P1_004N", # Black, not Hispanic or Latino
  year = 2020,
  sumfile = "pl",
  summary_var = "P1_001N",
  geometry = TRUE
  )
  
  saveRDS(black_nd_20, file=here("data", "black_nd_20.RDS"))
}
```

```{r loading state county datasets, echo = FALSE}
if (!file.exists(here("data", "ca_county_20.RDS"))) {
  ca_county_data <- get_acs(
  geography = "county",
  variables = c(
    total_households = "H1_001N",
    vacant_households = "H1_003N",
    median_household_income = "B19013_001",
    black_pop = "B02001_003"
  ),
  summary_var = "B01003_001",
  state = "CA",
  year = 2020,
  output = "wide"
)
  
  saveRDS(ca_county_data, file=here("data", "ca_county_20.RDS"))
}

if (!file.exists(here("data", "fl_county_20.RDS"))) {
  fl_county_data <- get_acs(
  geography = "county",
  variables = c(
    total_households = "H1_001N",
    vacant_households = "H1_003N",
    median_household_income = "B19013_001",
    black_pop = "B02001_003"
  ),
  summary_var = "B01003_001",
  state = "FL",
  year = 2020,
  output = "wide"
)
  
  saveRDS(fl_county_data, file=here("data", "fl_county_20.RDS"))
}

if (!file.exists(here("data", "pa_county_20.RDS"))) {
  pa_county_data <- get_acs(
  geography = "county",
  variables = c(
    total_households = "H1_001N",
    vacant_households = "H1_003N",
    median_household_income = "B19013_001",
    black_pop = "B02001_003"
  ),
  summary_var = "B01003_001",
  state = "PA",
  year = 2020,
  output = "wide"
)
  
  saveRDS(pa_county_data, file=here("data", "pa_county_20.RDS"))
}
```

```{r load Black population data, echo = FALSE}
years <- lst(2016, 2017, 2018, 2019, 2020)
my_vars <- c(
   black_pop = "B02001_003"
)

# Learned a way to use map_dfr from this informative tutorial: https://mattherman.info/blog/tidycensus-mult-year/

if (!file.exists(here("data", "black_pop_5_yr.RDS"))) {
  black_pop_5yr <- map_dfr(
    years,
    ~ get_acs(
      geography = "state",
      variables = my_vars,
      year = .x,
      survey = "acs5",
      geometry = FALSE
    ),
    .id = "year"
  ) %>%
    select(-moe) %>%
    arrange(variable, NAME)
  
  saveRDS(black_pop_5yr, file=here("data", "black_pop_5_yr.RDS"))
}
```

```{r loading median household income for Black, White, Asian race alone}
if (!file.exists(here("data", "medinc_race.RDS"))) {
  med_inc_race <- get_acs(
    geography = "state",
    variables = c(
      black = "B19013B_001",
      white = "B19013A_001",
      asian = "B19013D_001"
    ),
    geometry = FALSE,
    year = 2020,
    output = "wide"
  )
  
  saveRDS(med_inc_race, file=here("data", "medinc_race.RDS"))
}

if (!file.exists(here("data", "medinc_race_long.RDS"))) {
  med_inc_race <- get_acs(
    geography = "state",
    variables = c(
      black = "B19013B_001",
      white = "B19013A_001",
      asian = "B19013D_001"
    ),
    geometry = FALSE,
    year = 2020
  )
  
  saveRDS(med_inc_race, file=here("data", "medinc_race_long.RDS"))
}
```

```{r read in data from RDS files, echo = FALSE}
black_10 <- as_tibble(readRDS(here("data", "black_pop_10.RDS")))
black_20 <- as_tibble(readRDS(here("data", "black_pop_20.RDS")))
total_pop_10 <- as_tibble(readRDS(here("data", "total_pop_10.RDS")))
total_pop_20 <- as_tibble(readRDS(here("data", "total_pop_20.RDS")))
black_dc_20 <- as_tibble(readRDS(here("data", "black_dc_20.RDS")))
black_nd_20 <- as_tibble(readRDS(here("data", "black_nd_20.RDS")))

ca_county <- as_tibble(readRDS(here("data", "ca_county_20.RDS")))
fl_county <- as_tibble(readRDS(here("data", "fl_county_20.RDS")))
pa_county <- as_tibble(readRDS(here("data", "pa_county_20.RDS")))
black_pop_5yr <- as_tibble(readRDS(here("data", "black_pop_5_yr.RDS")))
med_inc_race <- as_tibble(readRDS(here("data", "medinc_race.RDS")))
med_inc_race_long <- as_tibble(readRDS(here("data", "medinc_race_long.RDS")))
```

# About

The purpose of this dashboard is to visualize US Census Data through two lenses: looking at the Black population data across the states, and state/county level data such as median household income across different dimensions of race.

The original data was obtained from the US decennial census data, as well as the American Community Survey, both from the Census Bureau.

Here's a link to the original website of the United States Census Bureau: https://www.census.gov/en.html

And a link to the American Community Survey (ACS): https://www.census.gov/programs-surveys/acs

Additionally, the data downloaded for this dashboard can be found in the following GitHub directory: https://github.com/jhu-statprogramming-fall-2023/biostat777-project4-tamisha-s/tree/ba9b332a7ff6a43c934b77c4837631adf7ec1aa1/data

# The Data

The `tidycensus` homepage describes the basic usage of the data, as well as shows a link to the data table/understanding of the variables.

Link: https://walker-data.com/tidycensus/articles/basic-usage.html

# Median Income by County

## Column {.sidebar}

Visualizing the median household income in $ by county in three states: California (CA), Florida (FL), and Pennsylvania (PA)

## Row {.tabset .tabset-fade}

### Household income by county in CA (top 10)

```{r CA median household income top 10}
ca_county %>%
  mutate(NAME = gsub(" County, California", "", NAME)) %>%
  top_n(10, median_household_incomeE) %>%
  ggplot(aes(x = median_household_incomeE, y = reorder(NAME, median_household_incomeE))) +
  geom_errorbarh(aes(xmin = median_household_incomeE - median_household_incomeM, xmax = median_household_incomeE + median_household_incomeM)) +
  geom_point(color = "purple", size = 3) +
  labs(
    title = "Household income ($) by county in California (top 10)",
    subtitle = "2016-2020 American Community Survey",
    y = "",
    x = "ACS estimate (bars represent margin of error)"
  )
```

### Household income by county in CA (bottom 5)

```{r CA median household income bottom 5}
ca_county %>%
  mutate(NAME = gsub(" County, California", "", NAME)) %>%
  top_n(-5, median_household_incomeE) %>%
  ggplot(aes(x = median_household_incomeE, y = reorder(NAME, -median_household_incomeE))) +
  geom_errorbarh(aes(xmin = median_household_incomeE - median_household_incomeM, xmax = median_household_incomeE + median_household_incomeM)) +
  geom_point(color = "red", size = 3) +
  labs(
    title = "Household income ($) by county in California (bottom 5)",
    subtitle = "2016-2020 American Community Survey",
    y = "",
    x = "ACS estimate (bars represent margin of error)"
  )
```

## Row {.tabset .tabset-fade}

### Household income by county in FL (top 10)

```{r FL median household income top 10}
fl_county %>%
  mutate(NAME = gsub(" County, Florida", "", NAME)) %>%
  top_n(10, median_household_incomeE) %>%
  ggplot(aes(x = median_household_incomeE, y = reorder(NAME, median_household_incomeE))) +
  geom_errorbarh(aes(xmin = median_household_incomeE - median_household_incomeM, xmax = median_household_incomeE + median_household_incomeM)) +
  geom_point(color = "purple", size = 3) +
  labs(
    title = "Household income ($) by county in Florida (top 10)",
    subtitle = "2016-2020 American Community Survey",
    y = "",
    x = "ACS estimate (bars represent margin of error)"
  )
```

### Household income by county in FL (bottom 5)

```{r FL median household income bottom 5}
fl_county %>%
  mutate(NAME = gsub(" County, Florida", "", NAME)) %>%
  top_n(-5, median_household_incomeE) %>%
  ggplot(aes(x = median_household_incomeE, y = reorder(NAME, -median_household_incomeE))) +
  geom_errorbarh(aes(xmin = median_household_incomeE - median_household_incomeM, xmax = median_household_incomeE + median_household_incomeM)) +
  geom_point(color = "red", size = 3) +
  labs(
    title = "Household income ($) by county in Florida (bottom 5)",
    subtitle = "2016-2020 American Community Survey",
    y = "",
    x = "ACS estimate (bars represent margin of error)"
  )
```

## Row {.tabset .tabset-fade}

### Household income by county in PA (top 10)

```{r PA median household income top 10}
pa_county %>%
  mutate(NAME = gsub(" County, Pennsylvania", "", NAME)) %>%
  top_n(10, median_household_incomeE) %>%
  ggplot(aes(x = median_household_incomeE, y = reorder(NAME, median_household_incomeE))) +
  geom_errorbarh(aes(xmin = median_household_incomeE - median_household_incomeM, xmax = median_household_incomeE + median_household_incomeM)) +
  geom_point(color = "purple", size = 3) +
  labs(
    title = "Household income ($) by county in Pennsylvania (top 10)",
    subtitle = "2016-2020 American Community Survey",
    y = "",
    x = "ACS estimate (bars represent margin of error)"
  )
```

### Household income by county in PA (bottom 5)

```{r PA median household income bottom 5}
pa_county %>%
  mutate(NAME = gsub(" County, Pennsylvania", "", NAME)) %>%
  top_n(-5, median_household_incomeE) %>%
  ggplot(aes(x = median_household_incomeE, y = reorder(NAME, -median_household_incomeE))) +
  geom_errorbarh(aes(xmin = median_household_incomeE - median_household_incomeM, xmax = median_household_incomeE + median_household_incomeM)) +
  geom_point(color = "red", size = 3) +
  labs(
    title = "Household income ($) by county in Pennsylvania (bottom 5)",
    subtitle = "2016-2020 American Community Survey",
    y = "",
    x = "ACS estimate (bars represent margin of error)"
  )
```

# Median Income by Race

## Row {.sidebar}

Visualizing the median household income (in \$) by race for three states - California, Florida, and Pennsylvania

## Row

### Household income by race for Black, White, and Asian for all the states

```{r}
med_inc_wrangled <- med_inc_race %>%
  rename(black = blackE) %>%
  rename(white = whiteE) %>%
  rename(asian = asianE) %>%
  gather(key = race, value = medinc, c(black, white, asian)) %>%
  select(NAME, race, medinc)

med_inc_wrangled %>%
  ggplot(aes(fill = race, y = medinc, x = NAME)) +
  geom_bar(position = "dodge", stat = "identity") +
  labs(
    x = "State",
    y = "Household Income in $$"
  ) +
  theme(axis.text.x = element_text(angle=50, vjust=0.5, size=5))
```

# Black Population Trend

## Column {.sidebar}

Visualizing Black population data over the last 5 years of the ACS data

## Column

### Black population trends over 5 years grouped by state

```{r}
black_pop_5yr %>%
  mutate_if(is.character, as.factor) %>%
  group_by(year) %>%
  ggplot(aes(x = year, y = estimate, group = NAME)) + 
  geom_line() +
  geom_point() +
  facet_wrap(~ NAME) + 
  labs(
    subtitle = "2016-2020 American Community Survey",
    x = "Year",
    y = "# of people"
  )
```

# State Household Income

## Column {data-width = 750}

### Household Income by State

```{r}
med_inc_race_long %>%
  mutate_if(is.character, as.factor) %>%
  ggplot(aes(NAME, estimate, color=variable, group=variable)) +
  geom_line() +
  geom_point() +
  theme(axis.text.x = element_text(angle=50, vjust=0.5, size=5)) +
  labs(
    x = "State",
    y = "Median Household Income",
    color = "Race"
  )
```

## Column {data-width = 350}

### Zero In: California Data

```{r}
ca_county %>% 
  mutate(NAME = gsub(" County, California", "", NAME)) %>%
  mutate(percent_black = format(round(100 * (black_popE / summary_est), 2), nsmall = 2)) %>%
  select(NAME, median_household_incomeE, percent_black) %>%
  rename(county = NAME, "median household income" = median_household_incomeE, "percent black" = percent_black) %>%
  arrange(desc("percent black")) %>% 
  DT::datatable()
```

# Analysis

## Row

### Plotting total population by state for 2010

```{r plotting total population by state for 2010}
total_pop_10 %>% 
  arrange(desc(value)) %>%
  mutate(ten_thou_pop = map_dbl(value, function(x) { return(x/10000) }) ) %>%
  ggplot(aes(x = ten_thou_pop, y = reorder(NAME, ten_thou_pop))) +
  geom_point() +
  labs(
    title = "Total population by state from the 2010 US Census",
    subtitle = "Arranged in descending order of population",
    x = "Population (in ten thousands)",
    y = "US State"
  )
```

### Plotting total population by state for 2020

```{r plotting total population by state for 2020}
total_pop_20 %>% 
  arrange(desc(value)) %>%
  mutate(ten_thou_pop = map_dbl(value, function(x) { return(x/10000) }) ) %>%
  ggplot(aes(x = ten_thou_pop, y = reorder(NAME, ten_thou_pop))) +
  geom_point() +
  labs(
    title = "Total population by state from the 2020 US Census",
    subtitle = "Arranged in descending order of population",
    x = "Population (in ten thousands)",
    y = "US State"
  )
```

### Visualizing Black population as percentage of total population in 2020

```{r visualizing Black population as percentage of total population in 2020}
p <- st_sf(black_20) %>%
  mutate(percent = 100 * (value / summary_value)) %>%
  ggplot(aes(fill = percent)) +
  geom_sf() +
  scale_fill_distiller(palette = "RdPu", 
                       direction = 1) +
  labs(title = "Black population by State, 2020",
       caption = "Data source: 2020 1-year ACS, US Census Bureau",
       fill = "Percentage of population that is Black") + 
  theme_void()

p
```

### Visualizing Black population in DC

```{r visualize Black populations in DC}
dc_black_pop_20 <- st_sf(black_dc_20) %>%
  mutate(percent = 100 * (value / summary_value))
  
tm_shape(dc_black_pop_20) +
  tm_polygons(col = "percent",
          style = "quantile",
          n = 5,
          palette = "Purples",
          title = "2020 US Census") + 
  tm_layout(title = "Percent Black\nby Census tract in DC",
            frame = FALSE,
            legend.outside = TRUE)
```

## Analyzing change in Black population since 2010

```{r analyzing change in Black population since 2010}
black_10_clean <- black_10 %>%
  st_drop_geometry() %>%
  select(GEOID, value, summary_value) %>%
  rename(value10 = value) %>%
  rename(summary10 = summary_value)

black_20_clean <- black_20 %>%
  rename(value20 = value) %>%
  rename(summary20 = summary_value)

black_joined <- black_20_clean %>%
  left_join(black_10_clean, by = "GEOID")

black_change <- black_joined %>%
  mutate(
    total_change = value20 - value10,
    percent_change = 100 * (total_change / value10)
  )
```

### Visualizing percentage change for Black population

```{r visualizing percentage change for Black population}
black_change %>%
  ggplot(aes(x = percent_change, y = reorder(NAME, percent_change))) +
  geom_point() +
  labs(
    title = "Visualizing percentage change for Black population from the 2010 and 2020 US census data",
    subtitle = "Arranged in descending order of percentage change",
    x = "Percent change from 2010 to 2020",
    y = "State"
  )
```

### Visualizing Black population in ND

```{r}
nd_black_pop_20 <- st_sf(black_nd_20) %>%
  mutate(percent = 100 * (value / summary_value))
  
tm_shape(nd_black_pop_20) +
  tm_polygons(col = "percent",
          style = "quantile",
          n = 5,
          palette = "Purples",
          title = "2020 US Census") + 
  tm_layout(title = "Percent Black\nby Census tract in ND",
            frame = FALSE,
            legend.outside = TRUE)
```

### Compare state with highest percentage of Black population (DC) with state with most Black population growth (ND)

```{r facet DC vs ND}
black_pop_10 <- black_10 %>%
  mutate(year = 2010)

black_pop_20 <- black_20 %>%
  mutate(year = 2020)

black_dc_nd <- black_pop_10 %>%
  bind_rows(black_pop_20) %>%
  rename(state = "NAME") %>%
  mutate(percent_black = value * 100 /summary_value) %>%
  mutate(year = as.integer(year)) %>%
  mutate(state = fct(state)) %>%
  filter(state == "North Dakota" | state == "District of Columbia") %>%
  group_by(state)

black_dc_nd %>%
  ggplot(aes(year, percent_black, fill = state)) +
  geom_bar(stat = "identity") +
  geom_line() +
  geom_point() +
  facet_grid(rows = vars(state)) +
  labs(
    title = "Comparing highest Black population with most Black population growth",
    subtitle = "Looking at DC and ND from US Census Data 2010-2020",
    x = "Year",
    y = "Percentage of population that is Black"
  )
```
